Block bootstrap consistency\\under weak assumptions

### 2015-12-31 Paper.tex

  \noindent \textsc{jel} Classification: C12, C15

@@

\newpage
\noindent Block bootstraps, e.g.\ the Moving Blocks
\citep{Kun:89,LiS:92}, Circular Block \citep{PoR:92}, and Stationary
Bootstraps \citep{PoR:94}, have become popular in Economics, partly
because they do not require the researcher to make parametric
assumptions about the data generating process.  They are valid under
general weak dependence and moment conditions.  Some recent papers
(\citealp{GoW:02}; \citealp{GoJ:03}) relax the dependence and moment
conditions of the original papers to fit with those commonly used in
Econometrics based on Near-Epoch-Dependence (\ned).%
\footnote{\citet{GoW:02} show that these bootstrap methods can be
  applied to heterogeneous $L_{2+\delta}$-\ned\ processes of size
  $-2(r-1)/(r-2)$ on a strong mixing sequence of size
  $-r(2+\delta)/(r-2)$, where $r > 2$ and $\delta >0$, when the
  original series has uniformly bounded $3r$-moments.  \citet{GoJ:03}
  relax these conditions to $L_{2+\delta}$-\ned\ of size $-1$ and
  $r+\delta$ moments for the original series, and size
  $-(2+\delta)(r+\delta)/(r-2)$ for the underlying mixing series.
  Both papers require that the expected block length grow with $n$ and
  be $o(n^{1/2})$.  \cite{GoP:11} discuss these issues further.}%
\footnote{An array $\{X_{nt}\}$ is an $L_{\rho}$-\ned\ process on a
  mixing array $\{V_{nt}\}$ if
  \begin{equation}
    \| X_{nt} - \E(X_{nt}
    \mid V_{n,t-m},\dots,V_{n,t+m}) \|_{\rho} \leq d_{nt} v_m
  \end{equation}
  with $v_m \to 0$ as $m \to \infty$ and $\{d_{nt}\}$ an array of
  positive constants.  It is of size $-\gamma$ if $v_m = O(m^{-\gamma
    - \delta})$ for all $\delta>0$.  Dropping the index ``$n$'' gives
  the series definition.  Note that strong and uniform mixing arrays
  are not required to be stationary.} %
But these conditions are still stronger than
required for a \clt\ to hold; \citet{Jon:97} has established the \clt\
under $L_2$-\ned\ with smaller size and moment
restrictions.%
\footnote{\citet{Jon:97} proves that the \clt\ holds for averages of
  $L_2$-\ned\ processes of size $-1/2$ on a strong mixing series of
  size $-r/(r-2)$, $r > 2$ and the original series having bounded
  $r$-moments.} %
This paper shows that these block bootstrap
methods consistently estimate the distribution of the sample mean
under \citepos{Jon:97} assumptions, and show that an \fclt\ holds as
well.%
\footnote{\citet{Rad:96} proves consistency for the Moving Blocks
  Bootstrap for any stationary strong mixing sequence that satisfies
  the \clt. This paper uses a similar method of proof to his, but also
  accommodates nonstationary sequences and the Stationary
  Bootstrap.} %
It also relaxes \citepos{GoW:02} and \citepos{GoJ:03} requirement that
the expected block length be $o(n^{1/2})$ to the original papers'
requirement that it be $o(n)$.

The proof exploits the conditional independence of the blocks in each
bootstrap.  Each bootstrap proceeds by drawing blocks of $M$
consecutive observations from the original time series, and then
pasting these blocks together to create the new bootstrap time series.
The Moving Blocks bootstrap does exactly that; the Circular Block
bootstrap ``wraps'' the observations, so that $(X_{n-1}, X_n, X_1,
X_2)$, for example, is a possible block of length four (letting $X_t$
denote the original time series).  The Stationary Bootstrap wraps the
observations and also draws $M$ at random for each block;
\citet{PoR:94} suggest drawing $M$ from the geometric distribution.
As the name suggests, the series produced by the Stationary Bootstrap
are strictly stationary, while those produced by the other methods are
not.  Although the Stationary Bootstrap was believed to be much less
efficient than other block bootstrap methods due to results of
\citet{Lah:99}, \citet{Nor:09} has shown that it is only slightly less
efficient than the other block-bootstrap methods discussed in this
paper, and has efficiency identical to that of the non-overlapping
block bootstrap.  Consequently, there has been renewed interest in the
Stationary Bootstrap since stationarity of the bootstrap samples can
be a useful property in theoretical research.  \cite{KrP:11} provides
a recent review of the bootstrap for time-series
processes,%
\footnote{Also see the discussion papers:
  \citet{Dah:11,GoP:11,Hor:11,JeM:11,KrP:11b}.} %
and \citet{GoP:11} further discuss recent developments in
block-bootstraps.

Theorem~\ref{main-bootstrap-clt} presents the main result, asymptotic normality of
the distribution of bootstrapped partial sums.  This paper adopts the
standard notation that $\E^{*}$, $\var^{*}$, etc.\ are the usual
operators with respect to the probability measure induced by the
bootstrap and will use explicit stochastic array notation
for precision.  Also note that all results are presented for the
scalar case but generalize immediately to random vectors.  All of the
proofs are presented in the appendix; only proofs for the Stationary
Bootstrap are presented, since proofs for the other methods are
similar and easier to follow.  All limits are taken as $n \to \infty$
unless otherwise noted and $\lVert \cdot \rVert_r$ denotes the
$L_r$-norm.

\begin{thm}\label{main-bootstrap-clt}
  Suppose the following conditions hold.
  \begin{enumerate}
  \item $X_{nt}$ is $L_2$-\ned\ of size $-1/2$ on an array
    $V_{nt}$ that is either strong mixing of size $-r/(r-2)$ or
    uniform mixing of size $-r/2(r-1)$, with $r > 2$.  The
    \ned\ magnitude indices are denoted $\{d_{nt}\}$.
  \item The array $\mu_{nt} - \bar \mu_n$ is uniformly bounded
    and $\sum_{t=1}^n (\mu_{nt} - \bar \mu_n)^2 \to 0$,
    where $\E X_{nt} = \mu_{nt}$ and $\bar{\mu}_n = n^{-1} \sum_{t=1}^n
    \mu_{nt}$. Moreover, $\sqrt{n} \| \bar{X}_{n} - \bar\mu_n \|_2
    \to \sigma > 0$, with $\bar X_n = (1/n) \sum_{t=1}^n X_{nt}$.
  \item There exists an array of positive real numbers $\{c_{nt}\}$
    such that $(X_{nt} - \mu_{nt})/c_{nt}$ is uniformly $L_r$-bounded
    and $c_{nt}$ and $d_{nt}/c_{nt}$ are uniformly bounded in $n$
    and~$t$.
  \item $X_{nt}^{*}$ is generated by the Stationary Bootstrap with
    geometric block lengths with success probability $p_n$, $p_n = c
    n^{-a}$ and $a,c \in (0,1)$, or by the Moving or Circular Block
    bootstrap with block length $M_n$ such that $M_n \sim n^a$ for
    $a \in (0,1)$.  Let $M_{ni}$ be the block length of the $i$th
    block, $i=1,\dots,J_n$, and define $K_{n0} = 0$ and $K_{nj} =
    \sum_{i=1}^j M_{ni}$. The last block, $M_{n,J_n}$, is defined as
    $K_{n,J_n} - K_{n,J_n-1}$, so $K_{n,J_n} = n$ a.s.
  \end{enumerate}
  Then $\hat\sigma^{*2} \to^p \sigma^2$,
  \begin{equation}\label{eq:36}
    \sup_x \big\lvert \pr^{*}\big[\sqrt{n}(\bar X_{n}^{*} - \E^* \bar X_n^*) \leq x \big]
    - \pr \big[\sqrt{n}(\bar X_{n} - \E \bar X_n) \leq x \big] \big\rvert \to^p 0,
  \end{equation}
  and
  \begin{equation}
    \label{eq:2}
    \sup_x \big\lvert \pr^{*}\big[
    \sqrt{n}(\bar X_{n}^{*} -  \E^* \bar X_n^*) \big/ \hat\sigma_n^{*}
    \leq x \big] - \Phi(x) \big\rvert \to^p 0,
  \end{equation}
  where $\Phi$ is the \textsc{cdf} of the Standard Normal
  distribution and
    \begin{equation}
      \label{eq:3}
      \hat{\sigma}_n^{*2} = \tfrac{1}{n} \sum_{j=1}^{J_n}
      \Big\{\sum_{t=K_{n,j-1}+1}^{K_{nj}} (X_{nt}^{*} - \bar X_{n}^{*})\Big\}^2.
    \end{equation}
\end{thm}

@@

Theorem~\ref{main-bootstrap-clt} relies on a general insight about the
variance of the sample mean under the bootstrap-induced
distribution. It is well-known that a key step in proving the \clt\
for arbitrary dependent processes is demonstrating that the squared
elements converge to a positive and finite limit; i.e.\ if
$\{Z_{nj};~j=1,\dots,J_n\}$ is a representative stochastic array,
$\sum_j Z_{nj}^2 \to^p \sigma^2$ is an important necessary condition
for $\sum_j Z_{nj} \to^d N(0,\sigma^2)$.  (See Section 3.2 of
\citealp{HaH:80}, for further discussion.)  For martingale difference
arrays, each $Z_{nj}$ is one of the original random variables $X_{nt}$
(typically normalized by $1/\sqrt{n}$), but for other forms of
dependence (\ned\ or mixingale arrays, for example, as in
\citealp{Jon:97}) each $Z_{nj}$ is a contiguous block of the original
random variables,
\begin{equation*}
  Z_{nj} = \tfrac{1}{\sqrt{n}} \sum_{t=(j-1) M_n +1}^{j M_n} (X_{nt} - \mu_{nt}),
\end{equation*}
that adds up to the original summation, plus potentially a negligible
residual, so $\sum_j Z_{nj}= (1/\sqrt{n}) \sum_t (X_{nt} - \mu_{nt}) +
o_p(1)$. In \citet{Jon:97}, for example, the \clt\ for mixingale
arrays assumes that there exists a sequence of blocks such that $\sum_j
Z_{nj}^2$ converges i.p., and the \ned\ \clt\ establishes conditions
under which such a $Z_{nj}$ exists.

Our insight is that the expectation of squared blocks of the bootstrap
process can be expressed as a sequence of \emph{contiguous} blocks of
the original process, so the arguments that establish convergence of
the original squared blocks can be applied with only minor changes to
the bootstrapped blocks. Consider the Moving Blocks Bootstrap,%
\footnote{To make this presentation as simple as possible, assume for
  now that $n = M_n J_n$ exactly.} %
for example, and let
\begin{equation*}
  Z_{nj}^* = \tfrac{1}{\sqrt{n}} \sum_{t=(j-1) M_n + 1}^{j M_n} (X_{nt}^* - \bar X_n).
\end{equation*}
Then, conditional on the data, the $Z_{nj}^{*2}$ are independent can
be expected to obey an \lln, so %
$\sum_j (Z_{nj}^{*2} - \E^* Z_{nj}^{*2}) \to^p 0$ %
and the CLT for the bootstrapped array requires $\sum_j \E^*
Z_{nj}^{*2}$ to converge to a positive and finite limit.  But, since
$\E^*$ only averages over the starting point of each block, we have
\begin{align*}
  \E^* Z_{nj}^{*2} &=
  \tfrac{1}{n} \sum_{\tau = 0}^{n-M_n}
  \Big(\tfrac{1}{\sqrt{n}}
  \sum_{t= \tau + 1}^{\tau + M_n} (X_{nt} - \bar X_n) \Big)^2 \\
  &= \tfrac{1}{n} \sum_{\tau_0 = 0}^{M_n-1}
  \sum_{j=1}^{J_n-1}
  \Big(\tfrac{1}{\sqrt{n}}
  \sum_{t= (j - 1) M_n + \tau_0 + 1}^{j M_n + \tau_0} (X_{nt} - \bar X_n) \Big)^2
\end{align*}
after grouping blocks separated by $M_n$ periods. For each $\tau_0$,
the summation can be expected to converge in probability through the
same arguments that were used to establish the \clt\ for the original
array.%
\footnote{Some of the details of the argument will typically need to
  change, since the original \clt\ only requires convergence for
  $\tau_0 = 0$, but these details are often incidental to the original
  argument.}%
\footnote{Lemmas~\ref{bootstrap-variance-convergence-lemma} and
  \ref{bootstrapped-zsq-uniform-integrability-lemma} are particularly
  strong demonstrations of this argument.} %
A similar representation is available for the Circular and Stationary
bootstraps.

In short, the basic approach that we use to prove Theorem~\ref{main-bootstrap-clt}
is based on a fundamental connection between the second moments of the
bootstrap process and the sum of squared blocks of the original
array. Even though the details of our proof rely on specific
techniques for \ned\ arrays, this connection implies that block
bootstraps are typically consistent when the original dependent array
obeys the \clt, and the connection should be useful for proving
consistency of the bootstrap under other dependence conditions.

For some uses, a \clt\ is inadequate and an \fclt\ is necessary.
Theorem~\ref{main-bootstrap-clt} can be extended to give an \fclt\
(using arguments from \citealp{JoD:00b}). We show in
Theorem~\ref{main-bootstrap-invariance-theorem} that the partial sum
of the bootstrapped process obeys an \fclt\ and can be used to derive
critical values for other test statistics under the same assumptions
as Theorem~\ref{main-bootstrap-clt}.

@@

For both results, we only present a proof for the stationary
bootstrap. The moving blocks and circular block bootstrap follow the
same general argument but are simpler. First we will define some
notation that will be used repeatedly in our results.

@@

\subsection*{Proof of Theorem~\ref{main-bootstrap-clt}}
{%
\newcommand{\Db}[1][j]{D_{n#1}^{*}}
\newcommand{\Zsum}{\sum_{j=1}^{J_n} \Zb / \sigma_n}
First we prove that
\begin{equation}\label{eq:37}
  \sup_x \big\lvert \pr^{*}\big[
  \sqrt{n}(\bar X_{n}^{*} -  \bar X_n) \big/ \sigma_n^{*}
  \leq x \big] - \Phi(x) \big\rvert \to^p 0
\end{equation}
where $\sigma_n^{*2} = n \E^* (\bar X_n^* - \bar X_n)^2$.
Rewrite $\sqrt{n}(\bar X_{n}^{*} - \bar X_n)$ as in
Equation~\eqref{eq:16}, so
\begin{equation*}
  \tfrac{1}{\sqrt{n}} \sum_{t=1}^n (X_{nt}^* - \bar X_n) =
  \sum_{j=1}^{J_n} Z_{nj}^*
\end{equation*}
and $\{\Zb / \sigma_n^*, \Gs_{nj}\}$ is a martingale difference
array. Moreover,
\begin{equation}
  \label{eq:18}
  \Pm^{*} \Big[\sum_{j=1}^{J_n} \Zb / \sigma_n^* \leq x \Big] - \Phi(x) \to^p 0,
\end{equation}
for all $x$ if $\sigma_n^{*2} \to^p \sigma^2$ (which ensures that
$\sigma_n^{*2}$ is uniformly a.s.\ positive and holds by
Lemma~\ref{bootstrap-variance-convergence-lemma}) and the following
two conditions hold for all positive $\epsilon$:
\begin{equation}
  \label{eq:19}
  \sum_{j=1}^{J_n} \Em^{*} \big(Z_{nj}^{*2} \ind\{Z_{nj}^{*2}  >
  \epsilon\}\big) \to^p 0
\end{equation}
and
\begin{equation}
  \label{eq:20}
  \Pm^{*}\Big[\ \Big|\sum_{j=1}^{J_n} Z_{nj}^{*2} - \sigma_n^{*2}
  \Big|\ > \epsilon \Big] \to^p 0
\end{equation}
since~\eqref{eq:19} and~\eqref{eq:20} ensure that $\Zb/\sigma_n^*$ obeys
a martingale difference \clt\ \citep[e.g.][Theorem 3.3]{HaH:80}.%
\footnote{Conditional on $X_{n1},\dots,X_{nn}$, $J_n$, and
  $M_{n1},\dots,M_{n,J_n}$, the only stochastic components of $\Zsum$
  are the start periods of each block, which are
  $\discreteuniform(1,\dots,n)$ and are independent of all of the
  other random variables in the information set used for conditioning.
  Consequently, $\Pm^*$ is a regular conditional probability and
  arguments like Hall and Heyde's (1980) Theorem 3.3
  apply without modification on this probability measure. See
  also Section 23.2 of \citet{Vaa:00}.

  Also note that Hall and Heyde's Theorem 3.3 as stated imposes an
  additional restriction on the sigma-fields. However, as Hall and
  Heyde discuss on pages 59 and 63--64, that condition is unnecessary
  here because the $\sigma_n^{*2}$ is measurable with respect to
  all of the $\Gs_{nj}$.} %

For~\eqref{eq:20}, since $Z_{nj}^*$ and $Z_{nk}^*$ ($k \neq j$) are uncorrelated
conditional on $X_{n1},\dots,X_{nn},$ and $\Ms$, we have
\begin{equation*}
  \sum_{j=1}^{J_n} Z_{nj}^{*2} - \sigma_n^{*2} =
  \sum_{j=1}^{J_n} \Big( Z_{nj}^{*2} - (1/J_n) \E^* \sum_{j=1}^{J_n} Z_{nj}^{*2} \Big)
\end{equation*}
almost surely. But
\begin{equation*}
   \Big\{\tfrac{n}{M_{nj}} \Big(Z_{nj}^{*2} - (1/J_n) \E^* \sum_{j=1}^{J_n} Z_{nj}^{*2}\Big),
   \ \Gs_{nj} \Big\}
\end{equation*}
is a uniformly-integrable martingale difference array
by Lemma~\ref{bootstrapped-zsq-uniform-integrability-lemma},
and satisfies the \lln\ (e.g., \citealp{Dav:94}, Theorem 19.7), so
this sum converges to zero in conditional probability, proving~\eqref{eq:20}.

>>>>

  \noindent \allcaps{jel} Classification: C12, C15

@@

\newpage
\noindent Block bootstraps, e.g.\ the Moving Blocks
(\citealp{Kun:89}, and \citealp{LiS:92}), Circular Block \citep{PoR:92}, and Stationary
Bootstraps \citep{PoR:94}, have become popular in Economics, partly
because they do not require the researcher to make parametric
assumptions about the data generating process.  They are valid under
general weak dependence and moment conditions.  Some recent papers
(\citealp{GoW:02}, and \citealp{GoJ:03}) relax the dependence and moment
conditions of the original papers to fit with the Near-Epoch-Dependence
(\ned) assumptions commonly used in econometrics.%
\footnote{\citet{GoW:02} show that these bootstrap methods can be
  applied to heterogeneous $L_{2+\delta}$-\ned\ processes of size
  $-2(r-1)/(r-2)$ on a strong mixing sequence of size
  $-r(2+\delta)/(r-2)$, where $r > 2$ and $\delta >0$, when the
  original series has uniformly bounded $3r$-moments.  \citet{GoJ:03}
  relax these conditions to $L_{2+\delta}$-\ned\ of size $-1$ and
  $r+\delta$ moments for the original series, and size
  $-(2+\delta)(r+\delta)/(r-2)$ for the underlying mixing series.
  Both papers require that the expected block length grow with $n$ and
  be $o(n^{1/2})$.  \cite{GoP:11} discuss these issues further.}%
\footnote{An array $\{X_{nt}\}$ is an $L_{\rho}$-\ned\ process on a
  mixing array $\{V_{nt}\}$ if
  \begin{equation}
    \| X_{nt} - \E(X_{nt}
    \mid V_{n,t-m},\dots,V_{n,t+m}) \|_{\rho} \leq d_{nt} v_m
  \end{equation}
  with $v_m \to 0$ as $m \to \infty$ and $\{d_{nt}\}$ an array of
  positive constants.  It is of size $-\gamma$ if $v_m = O(m^{-\gamma
    - \delta})$ for all $\delta>0$.  Dropping the index ``$n$'' gives
  the series definition.  Note that the underlying strong and uniform mixing arrays
  are not required to be stationary.} %
But these conditions are still stronger than
required for a \clt\ to hold; \citet{Jon:97} has established the \clt\
under $L_2$-\ned\ with smaller size and moment
restrictions.%
\footnote{\citet{Jon:97} proves that the \clt\ holds for averages of
  $L_2$-\ned\ processes of size $-1/2$ on a strong mixing series of
  size $-r/(r-2)$, $r > 2$ and the original series having bounded
  $r$-moments.} %
This paper shows that these block bootstrap
methods consistently estimate the distribution of the sample mean
under \citepos{Jon:97} assumptions and show that an \fclt\ holds as
well.%
\footnote{\citet{Rad:96} proves consistency for the Moving Blocks
  Bootstrap for any stationary strong mixing sequence that satisfies
  the \clt. This paper uses a similar method of proof to his, but also
  accommodates nonstationary sequences and the Stationary
  Bootstrap.} %
It also relaxes \citepos{GoW:02} and \citepos{GoJ:03} requirement that
the expected block length be $o(n^{1/2})$ to the original papers'
requirement that it be $o(n)$.

The proof exploits the conditional independence of the blocks in each
bootstrap.  Each bootstrap proceeds by drawing blocks of $M$
consecutive observations from the original time series, and then
pasting these blocks together to create the new bootstrap time series.
The Moving Blocks bootstrap does exactly that; the Circular Block
bootstrap ``wraps'' the observations, so that $(X_{n-1}, X_n, X_1,
X_2)$, for example, is a possible block of length four (letting $X_t$
denote the original time series).  The Stationary Bootstrap wraps the
observations and also draws $M$ at random for each block;
\citet{PoR:94} suggest drawing $M$ from the geometric distribution.
As the name suggests, the series produced by the Stationary Bootstrap
are strictly stationary, while those produced by the other methods are
not.  Although the Stationary Bootstrap was believed to be much less
efficient than other block bootstrap methods due to results of
\citet{Lah:99}, \citet{Nor:09} has shown that it is only slightly less
efficient than the other block-bootstrap methods discussed in this
paper, and has efficiency identical to that of the non-overlapping
block bootstrap.  Consequently, there has been renewed interest in the
Stationary Bootstrap since stationarity of the bootstrap samples can
be a useful property in theoretical research.  \cite{KrP:11} provides
a recent review of the bootstrap for time-series
processes%
\footnote{Also see the discussion papers by \citet{Dah:11},
\citet{GoP:11}, \citet{Hor:11}, \citet{JeM:11}, and
\citet{KrP:11b}.} %
and \citet{GoP:11} further discuss recent developments in
block-bootstraps.

Theorem~\ref{main-bootstrap-clt} presents the main result, asymptotic normality of
the distribution of bootstrapped sums. This paper adopts the
standard notation that $\E^{*}$, $\var^{*}$, etc.\ are the usual
operators with respect to the probability measure induced by the
bootstrap and will use explicit stochastic array notation
for precision.  Also note that all results are presented for the
scalar case but generalize immediately to random vectors.  All of the
proofs are presented in the appendix; only proofs for the Stationary
Bootstrap are presented, since proofs for the other methods are
similar and easier to construct.  All limits are taken as $n \to \infty$
unless otherwise noted and $\lVert \cdot \rVert_r$ denotes the
$L_r$-norm.

\begin{thm}\label{main-bootstrap-clt}
  Suppose the following conditions hold.
  \begin{enumerate}
  \item $X_{nt}$ is $L_2$-\ned\ of size $-1/2$ on an array
    $\{V_{nt}\}$ that is either strong mixing of size $-r/(r-2)$ or
    uniform mixing of size $-r/2(r-1)$, with $r > 2$.  The
    \ned\ magnitude indices are denoted $\{d_{nt}\}$.
  \item The array $\mu_{nt} - \bar \mu_n$ is uniformly bounded
    and $\sum_{t=1}^n (\mu_{nt} - \bar \mu_n)^2 \to 0$,
    where $\E X_{nt} = \mu_{nt}$ and $\bar{\mu}_n = n^{-1} \sum_{t=1}^n
    \mu_{nt}$. Moreover, $\sqrt{n} \| \bar{X}_{n} - \bar\mu_n \|_2
    \to \sigma > 0$, with $\bar X_n = (1/n) \sum_{t=1}^n X_{nt}$.
  \item There exists an array of positive real numbers $\{c_{nt}\}$
    such that $(X_{nt} - \mu_{nt})/c_{nt}$ is uniformly $L_r$-bounded
    and $c_{nt}$ and $d_{nt}/c_{nt}$ are uniformly bounded in $n$
    and~$t$.
  \item $X_{nt}^{*}$ is generated by the Stationary Bootstrap with
    geometric block lengths with success probability $p_n$, $p_n = c
    n^{-a}$ and $a,c \in (0,1)$, or by the Moving or Circular Block
    bootstrap with block length $M_n$ such that $M_n \sim n^a$ for
    $a \in (0,1)$.  Let $M_{ni}$ be the block length of the $i$th
    block, $i=1,\dots,J_n$, and define $K_{n0} = 0$ and $K_{nj} =
    \sum_{i=1}^j M_{ni}$. The last block, $M_{n,J_n}$, is defined as
    $K_{n,J_n} - K_{n,J_n-1}$, so $K_{n,J_n} = n$ a.s.
  \end{enumerate}
  Then $\hat\sigma^{*2} \to^p \sigma^2$,
  \begin{equation}\label{eq:36}
    \sup_x \big\lvert \pr^{*}\big[\sqrt{n}(\bar X_{n}^{*} - \E^* \bar X_n^*) \leq x \big]
    - \pr \big[\sqrt{n}(\bar X_{n} - \E \bar X_n) \leq x \big] \big\rvert \to^p 0,
  \end{equation}
  and
  \begin{equation}
    \label{eq:2}
    \sup_x \big\lvert \pr^{*}\big[
    \sqrt{n}(\bar X_{n}^{*} -  \E^* \bar X_n^*) \big/ \hat\sigma_n^{*}
    \leq x \big] - \Phi(x) \big\rvert \to^p 0,
  \end{equation}
  where $\Phi$ is the \allcaps{cdf} of the Standard Normal
  distribution and
    \begin{equation}
      \label{eq:3}
      \hat{\sigma}_n^{*2} = \tfrac{1}{n} \sum_{j=1}^{J_n}
      \Big\{\sum_{t=K_{n,j-1}+1}^{K_{nj}} (X_{nt}^{*} - \bar X_{n}^{*})\Big\}^2.
    \end{equation}
\end{thm}

@@

Theorem~\ref{main-bootstrap-clt} relies on a general insight about the
variance of the sample mean under the bootstrap-induced
distribution. It is well-known that a key step in proving the \clt\
for arbitrary dependent processes is demonstrating that the squared
elements converge to a positive and finite limit; i.e.\ if
$\{Z_{nj};~j=1,\dots,J_n\}$ is a representative stochastic array,
$\sum_j Z_{nj}^2 \to^p \sigma^2$ is an important necessary condition
for $\sum_j Z_{nj} \to^d N(0,\sigma^2)$.  (See Section 3.2 of
\citealp{HaH:80}, for further discussion.)  For martingale difference
arrays, each $Z_{nj}$ is one of the original random variables $X_{nt}$
(typically normalized by $1/\sqrt{n}$), but for other forms of
dependence (\ned\ or mixingale arrays, for example, as in
\citealp{Jon:97}) each $Z_{nj}$ is a contiguous block of the original
random variables,
\begin{equation*}
  Z_{nj} = \tfrac{1}{\sqrt{n}} \sum_{t=(j-1) M_n +1}^{j M_n} (X_{nt} - \mu_{nt}),
\end{equation*}
that adds up to the original summation (plus potentially a negligible
residual) so $\sum_j Z_{nj}= (1/\sqrt{n}) \sum_t (X_{nt} - \mu_{nt}) +
o_p(1)$. In \citet{Jon:97}, for example, the \clt\ for mixingale
arrays assumes that there exists a sequence of blocks such that $\sum_j
Z_{nj}^2$ converges i.p., and the \ned\ \clt\ establishes conditions
under which such a $Z_{nj}$ exists.

Our insight is that the expectation of squared blocks of the bootstrap
process can be expressed as a sequence of \emph{contiguous} blocks of
the original process, so the arguments that establish convergence of
the original squared blocks can be applied with only minor changes to
the bootstrapped blocks. Consider the Moving Blocks Bootstrap,%
\footnote{To make this presentation as simple as possible, assume for
  now that $n = M_n J_n$ exactly.} %
for example, and let
\begin{equation*}
  Z_{nj}^* = \tfrac{1}{\sqrt{n}} \sum_{t=(j-1) M_n + 1}^{j M_n} (X_{nt}^* - \bar X_n).
\end{equation*}
Then, conditional on the data, the $Z_{nj}^{*2}$ are independent can
be expected to obey an \lln, so %
$\sum_j (Z_{nj}^{*2} - \E^* Z_{nj}^{*2}) \to^p 0$ %
and the CLT for the bootstrapped array requires $\sum_j \E^*
Z_{nj}^{*2}$ to converge to a positive and finite limit.  But, since
$\E^*$ only averages over the starting point of each block, we have
\begin{align*}
  \E^* Z_{nj}^{*2} &=
  \tfrac{1}{n} \sum_{\tau = 0}^{n-M_n}
  \Big(\tfrac{1}{\sqrt{n}}
  \sum_{t= \tau + 1}^{\tau + M_n} (X_{nt} - \bar X_n) \Big)^2 \\
  &= \tfrac{1}{n} \sum_{\tau_0 = 0}^{M_n-1}
  \sum_{j=1}^{J_n-1}
  \Big(\tfrac{1}{\sqrt{n}}
  \sum_{t= (j - 1) M_n + \tau_0 + 1}^{j M_n + \tau_0} (X_{nt} - \bar X_n) \Big)^2
\end{align*}
after grouping blocks separated by $M_n$ periods. For each $\tau_0$,
the summation can be expected to converge in probability through the
same arguments that were used to establish the \clt\ for the original
array.%
\footnote{Some of the details of the argument will typically need to
  change because the original \clt\ only requires convergence for
  $\tau_0 = 0$, but these details are often incidental to the original
  argument.}%
\footnote{Lemmas~\ref{bootstrap-variance-convergence-lemma} and
  \ref{bootstrapped-zsq-uniform-integrability-lemma} are particularly
  strong demonstrations of this argument.} %
A similar representation is available for the Circular and Stationary
bootstraps.

In short, the basic approach that we use to prove Theorem~\ref{main-bootstrap-clt}
is based on a fundamental connection between the second moments of the
bootstrap process and the sum of squared blocks of the original
array. Even though the details of our proof rely on specific
techniques for \ned\ arrays, this connection implies that block
bootstraps are typically consistent when the original dependent array
obeys the \clt\ and the connection should be useful for proving
consistency of the bootstrap under other dependence conditions.

Theorem~\ref{main-bootstrap-clt} can be further extended to give an
\fclt\ using arguments from \citet{JoD:00b}. We show in
Theorem~\ref{main-bootstrap-invariance-theorem} that the partial sum
of the bootstrapped process obeys an \fclt\ and can be used to derive
critical values for other test statistics under the same assumptions
as Theorem~\ref{main-bootstrap-clt}.

@@

For both results, we only present a proof for the stationary
bootstrap. The moving blocks and circular block bootstrap follow the
same general argument but are simpler. First we will define some
notation that will be used repeatedly.

@@

\subsection*{Proof of Theorem~\ref{main-bootstrap-clt}}
{%
\newcommand{\Db}[1][j]{D_{n#1}^{*}}
\newcommand{\Zsum}{\sum_{j=1}^{J_n} \Zb / \sigma_n}
First we prove that
\begin{equation}\label{eq:37}
  \sup_x \big\lvert \pr^{*}\big[
  \sqrt{n}(\bar X_{n}^{*} -  \bar X_n) \big/ \sigma_n^{*}
  \leq x \big] - \Phi(x) \big\rvert \to^p 0
\end{equation}
where $\sigma_n^{*2} = n \E^* (\bar X_n^* - \bar X_n)^2$.
Rewrite $\sqrt{n}(\bar X_{n}^{*} - \bar X_n)$ as in
Equation~\eqref{eq:16}, so
\begin{equation*}
  \tfrac{1}{\sqrt{n}} \sum_{t=1}^n (X_{nt}^* - \bar X_n) =
  \sum_{j=1}^{J_n} Z_{nj}^*
\end{equation*}
and $\{\Zb / \sigma_n^*, \Gs_{nj}\}$ is a martingale difference
array. Moreover,
\begin{equation}
  \label{eq:18}
  \Pm^{*} \Big[\sum_{j=1}^{J_n} \Zb / \sigma_n^* \leq x \Big] - \Phi(x) \to^p 0,
\end{equation}
for all $x$ if $\sigma_n^{*2} \to^p \sigma^2$ (which ensures that
$\sigma_n^{*2}$ is uniformly a.s.\ positive and holds by
Lemma~\ref{bootstrap-variance-convergence-lemma}) and the following
two conditions hold for all positive $\epsilon$:
\begin{equation}
  \label{eq:19}
  \sum_{j=1}^{J_n} \Em^{*} \big(Z_{nj}^{*2} \ind\{Z_{nj}^{*2}  >
  \epsilon\}\big) \to^p 0
\end{equation}
and
\begin{equation}
  \label{eq:20}
  \Pm^{*}\Big[\ \Big|\sum_{j=1}^{J_n} Z_{nj}^{*2} - \sigma_n^{*2}
  \Big|\ > \epsilon \Big] \to^p 0
\end{equation}
since~\eqref{eq:19} and~\eqref{eq:20} ensure that $\Zb/\sigma_n^*$ obeys
a martingale difference \clt\ \citep[e.g.][Theorem 3.3]{HaH:80}.%
\footnote{Conditional on $X_{n1},\dots,X_{nn}$, $J_n$, and
  $M_{n1},\dots,M_{n,J_n}$, the only stochastic components of $\Zsum$
  are the start periods of each block, which are
  $\discreteuniform(1,\dots,n)$ and are independent of all of the
  other random variables in the information set used for conditioning.
  Consequently, $\Pm^*$ is a regular conditional probability and
  arguments like Hall and Heyde's (1980) Theorem 3.3
  apply without modification on this probability measure. See
  also Section 23.2 of \citet{Vaa:00}.
  Also note that Hall and Heyde's Theorem 3.3 as stated imposes an
  additional restriction on the sigma-fields. However, as Hall and
  Heyde discuss on pages 59 and 63--64, that condition is unnecessary
  here because $\sigma_n^{*2}$ is measurable with respect to
  all of the $\Gs_{nj}$.} %

For~\eqref{eq:20}, $Z_{nj}^*$ and $Z_{nk}^*$ (when $k \neq j$) are
conditionally uncorrelated given $X_{n1},\dots,X_{nn},$ and $\Ms$, which
implies
\begin{equation*}
  \sum_{j=1}^{J_n} Z_{nj}^{*2} - \sigma_n^{*2} =
  \sum_{j=1}^{J_n} \Big( Z_{nj}^{*2} - (1/J_n) \E^* \sum_{j=1}^{J_n} Z_{nj}^{*2} \Big)
\end{equation*}
almost surely. But
\begin{equation*}
   \Big\{\tfrac{n}{M_{nj}} \Big(Z_{nj}^{*2} - (1/J_n) \E^* \sum_{j=1}^{J_n} Z_{nj}^{*2}\Big),
   \ \Gs_{nj} \Big\}
\end{equation*}
is a uniformly-integrable martingale difference array
by Lemma~\ref{bootstrapped-zsq-uniform-integrability-lemma}
and satisfies the \lln\ (e.g., \citealp{Dav:94}, Theorem 19.7). So
this sum converges to zero in conditional probability, proving~\eqref{eq:20}.
